---
title: "Conversion from long format"
author: "baptiste"
date: today
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(grid)
library(patchwork)
read_chunk('_utilities.R')
source("_utilities.R")
```


## Reading long-format T-matrix files

Some programs output T-matrices in long format, similar to the example below:

```{r, echo=FALSE}
d <- read.table('../utilities/smarties.txt')
names(d) <- c('s','sp','l','lp','m','mp','Tr','Ti')
head(d)
```

Note that one advantage of this format is that only non-zero entries need to be provided, similar to a sparse matrix definition. This is useful for EBCM-like analytical methods where particle symmetries dictate that some T-matrix elements be identically zero (plane of symmetry, axis of symmetry, etc.).


## Converting to wide format

The example script [convert_longformat.jl](../utilities/convert_longformat.jl) reads this plain text dataset, converts the indices $(l,m,s)$ to [combined indices](indexing.html) $(u,u')$, and exports the data in the standard `.tmat.h5` format, adding the required metadata.

```{.julia}
{{< include ../utilities/convert_longformat.jl >}}
```


